### 2013-11-07 src2/thesis.tex
update

\textbf{Keywords:} Data-parallelism, Functional Programming, Embedded languages, GPU

@@

@@

This thesis applies embedded language methodology to data-parallel programming of 
massively parallel hardware. An Embedded language is a language that is implemented 
as a library for some existing host language. In this case, the functional programming 
language, Haskell is used as host language.

Computers today are increasingly parallel. Each new generation of general purpose processors
(CPUs) and of graphics processors (GPUs), brings more parallelism. The question 
of how to program these computers most efficiently, both from a productivity and 
a performance point of view, arise. Explicit parallel programming is hard and automatic 
means of parallelisation, while giving a sequential programming model to the programmer, 
often misses the target when it comes to performance. {\bf \emph{source??}}

The embedded language approach to programming language implementation is beneficial 
since we are not entirely sure of how to program current and future highly parallel 
computers. Implementing an embedded language allows us more rapid prototyping of ideas. 
We can experiment with what abstractions to expose to the programmer, without much 
effort. Other benefits of embedded languages often mentioned are: making use 
of the host language's library functions, if they are general enough. The host language 
can be used as a powerful macro language for the embedded language. An embedded 
language also provide a more leveled learning curve to programmers already familiar 
with the host language. For these programmers using an embedded language is more like  
learning about a new library.

@@

Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate large portions of chip area to caches, branch 
prediction and Instruction level parallelism (ILP), GPUs put additional cores. 
This comes at a price of increased programmer effort. GPUs have programmer managed 
shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application. With Obsidian we try to find the right level of abstraction for 
programming GPUs. We try to strike a balance between low-level control of details 
that influence performance and high-level abstractions to increase productivity and 
make programs more terse. The end goal with Obsidian is a language for writing 
high performance GPU programs that compare well to code written by CUDA~\citet{CUDA}  
(a C-dialect for GPU programming) experts, provided that both programmers have 
similar GPU expertise. The Obsidian programmer would however do so in a more 
composable manner. Obsidian programs are compiled into CUDA.

Obsidian is geared towards the programmer who is interested in GPU hardware. Making 
good use of Obsidian requires that the programmer is willing to make an effort 
of learning about GPUs and how to program them efficiently. However, another case 
is a programmer who does not want to invest that effort but wants to write a high-level 
parallel program and have it run reasonably well on different parallel architectures. 
In this case a language with even higher level of abstraction is preferable. It then 
becomes the job of a more advanced compiler to turn that into code for the targetted 
platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
operations on arrays, such as {\tt map}, {\tt reduce} and {\tt scan}. The programmer 
has no power over the parallel strategy that should be used to implement these 
operations but instead rely on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
Haskell embedding of Intel ArBB that was released as an embedded language in C++. 
EmbArBB relies entirely on the ArBB system for code generation and optimisation.

@@

Before CUDA, programmers still tried (very ingeniously) to make use of GPUs for 
non-graphics applications via the graphics API. The desired computation was expressed 
in graphics terms and the result computed through an act of rendering. Of course, since the 
available abstractions targeted the graphics domain, this added extra
burden on the programmer. CUDA was NVIDIA's first response to this problem. With CUDA 
there is a dialect of C, extended with a parallel programming model suitable for GPUs.

@@

@@

@@

\noindent The kernel takes two input arrays, {\tt i1} and {\tt i2}; the result is stored 
in {\tt r}. Note how the kernel makes use of three variables, {\tt threadIdx.x}, 
{\tt blockIdx.x} and {\tt gridDim.x}. These variables identify a thread uniquely within 
a grid. Here they are used to obtain an index into the provided arrays. The dimensions of 
the blocks and of the grid is set by a controling program upon launch of the kernel. This 
controling program is run by the CPU of the computer housing the GPU. the kernel is launched 
using the following syntax:

@@

\noindent The line of code above launches the kernel on one block of 32 threads. The zero 
indicates that the kernel uses no shared memory. If the kernel did use shared memory, the 
amount needed should have been specified there. Figure~\ref{fig:CUDACODE1} shows the complete 
sequence of commands for a minimal example using the kernel specified above.

In Obsidian the same kernel (the local computation) is implemented as follows:

\noindent The {\tt vecAdd} program takes two {\tt SPull} input arrays and gives one 
as result. The Obsidian {\tt vecAdd} is more general than the CUDA version. From it 
we can generate CUDA kernels for any type that supports {\tt (+)}, is in the {\tt Num} class.

There are some important differences between CUDA and Obsidian programs. One is that with Obsidian 
we generate kernels for fixed size local computations. This is to avoid doing out-of-bounds check 
which would mean conditionals in the code and conditionals are problematic from a performance 
point of view. Another difference is that in Obsidian the programmer specifies the entire 
computation, including how it is mapped over the blocks of the GPU. This is done implicitly 
in CUDA, based on the indexing arithmetic used in the kernel and the number of blocks and 
threads provided at kernel launch. This means that the Obsidian programmer needs to 
write a program that performs this mapping:

\begin{small} 
\begin{Verbatim}[samepage=true]
vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
vecAddG i1 i2  =
  pConcat $
  fmap push $
  zipWith vecAdd
          (splitUp 32 i1)
          (splitUp 32 i2)  
\end{Verbatim}
\end{small}

\subsection{Writing efficient GPU Kernels} 
 \emph{now takes up both CUDA and Obsidian aspects at the same time, pref. with code examples}

The CUDA programming model reflects the hierarchical architecture of GPUs. The programmer 
writes what are called {\em kernels}; in CUDA jargon a kernel refers to a sequential C program 
that is parameterised on a thread's identity. The kernel program is executed in parallel across 
all the threads of a block and many such blocks of threads can execute in parallel on the 
available MPs. The programmer needs to decompose the algorithm and data in such a way that 
communication will be local to the blocks.

Below is a very simple CUDA example to give some indication of what a kernel and 
the code to launch that kernel looks like.

\begin{small}
\begin{Verbatim}[samepage=true]
__global__ void kernel(int32_t *i1, int32_t *i2, int32_t *r) {

This kernel takes two input arrays, {\tt i1} and {\tt i2}, and computes 
element wise sums into the result array, {\tt r}. The {\tt threadIdx} variable identifies 
a thread within a block. There is also a {\tt blockIdx} variable that identifies 
the block a thread belongs to and {\tt blockDim.x} specifies the number of threads per block.

In CUDA there is a notion of {\em host} and {\em device}, the device is the GPU and 
its memory (Global memory, usually a few Gigabytes). The host refers to the system housing 
the GPU. The host runs a control program that launches kernels on the GPU. First memory 
needs to be allocated in the device to hold the inputs and outputs. Then data is copied from 
the host's system memory into device memory.

@@

\pagebreak

The syntax for launching a kernel is:

\begin{small}
\begin{Verbatim}[samepage=true]
name<<<Blocks, Threads/Block, BytesShared>>>(arg1,..,argN);
\end{Verbatim}
\end{small}

\noindent In the example above one block of 32 threads, using no shared memory, is launched.

\begin{small}
\begin{Verbatim}[samepage=true]
__global__ void kernel2(int32_t *i, int32_t *r) {

@@

\noindent A portion of the input array is stored into shared memory. A 
barrier synchronisation is used to ensure that each threads within the block 
has stored its value into shared memory (only necessary if threads will 
communicate across warp boundaries). After computing on the data locally, 
it is written back into global memory.

>>>>

\textbf{Keywords:} Data-parallelism, Embedded languages, Functional Programming, Graphics Processing Units

@@

% -------------------------------------------------------------------------
%  INTRO SECTION OF INTRO CHAPTER
% -------------------------------------------------------------------------

@@

This thesis applies embedded language methodology to data-parallel programming of 
massively parallel hardware. An embedded language is a language that is implemented 
as a library for some existing host language. In this case, the functional programming 
language Haskell is used as host language.

Computers today are increasingly parallel. Each new generation of general purpose processors
(CPUs) and of graphics processors (GPUs), brings more parallelism. The question 
of how to program these computers most efficiently, both from a productivity and 
a performance point of view, arises. Explicit parallel programming is hard and automatic 
means of parallelisation, while giving a sequential programming model to the programmer, 
often misses the target when it comes to performance. {\bf \emph{source??}}

The embedded language approach to programming language implementation is beneficial, 
since we are not entirely sure of how to program current and future highly parallel 
computers. Implementing an embedded language allows us more rapid prototyping of ideas. 
We can experiment with what abstractions to expose to the programmer, without much 
effort. Other benefits of embedded languages often mentioned are: making use 
of the host language's library functions, if they are general enough. The host language 
can be used as a powerful macro language for the embedded language. An embedded 
language also provides a more leveled learning curve to programmers already familiar 
with the host language. For these programmers, using an embedded language is more like  
learning about a new library.

@@

Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate large portions of chip area to caches, branch 
prediction and instruction level parallelism (ILP), GPUs put additional cores. 
This comes at a price of increased programmer effort. GPUs have programmer managed 
shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application. With Obsidian, we try to find the right level of abstraction for 
programming GPUs. We try to strike a balance between low-level control of details 
that influence performance and high-level abstractions to increase productivity and 
make programs more terse. The end goal with Obsidian is a language for writing 
high performance GPU programs that compare well to code written by CUDA~\citet{CUDA}  
(a C-dialect for GPU programming) experts, provided that both programmers have 
similar GPU expertise. The Obsidian programmer would however do so in a more 
composable manner. Obsidian programs are compiled into CUDA.

Obsidian is geared towards the programmer who is interested in GPU hardware. Making 
good use of Obsidian requires that the programmer is willing to make an effort 
to learn about GPUs and how to program them efficiently. However, another case 
is a programmer who does not want to invest that effort but wants to write a high-level 
parallel program and have it run reasonably well on different parallel architectures. 
In this case, a language with even higher level of abstraction is preferable. It then 
becomes the job of a more advanced compiler to turn that into code for the targetted 
platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
operations on arrays, such as {\tt map}, {\tt reduce} and {\tt scan}. The programmer 
has no power over the parallel strategy that should be used to implement these 
operations but instead relies on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
Haskell embedding of Intel ArBB that was released as an embedded language in C++. 
EmbArBB relies entirely on the ArBB system for code generation and optimisation.

% -------------------------------------------------------------------------
% Obsidian !
% -------------------------------------------------------------------------

@@

Before CUDA, programmers still tried (very ingeniously) to make use of GPUs for 
non-graphics applications via the graphics API. The desired computation was expressed 
in graphics terms and the result computed through an act of rendering. Of course, since the 
available abstractions targeted the graphics domain, this added extra
burden on the programmer. CUDA was NVIDIA's first response to this problem. CUDA 
is a dialect of C, extended with a parallel programming model suitable for GPUs.

@@

% -------------------------------------------------------------------------
% GPU Hardware 
% -------------------------------------------------------------------------

@@

% -------------------------------------------------------------------------
%  GPU Programming
% -------------------------------------------------------------------------

@@

\noindent The kernel takes two input arrays, {\tt i1} and {\tt i2}; the result is stored 
in {\tt r}. Note how the kernel makes use of three variables, {\tt threadIdx.x}, 
{\tt blockIdx.x} and {\tt gridDim.x}. These variables identify a thread uniquely within 
a grid. Here they are used to obtain an index into the provided arrays. The dimensions of 
the blocks and of the grid are set by a controling program upon launch of the kernel. This 
controling program is run by the CPU of the computer housing the GPU. The kernel is launched 
using the following syntax:

@@

\noindent This line of code launches the kernel on one block of 32 threads. The zero 
indicates that the kernel uses no shared memory. If the kernel did use shared memory, the 
amount needed should have been specified there. Figure~\ref{fig:CUDACODE1} shows the complete 
sequence of commands for a minimal example using the kernel specified above.

% -------------------------------------------------------------------------
% CUDACODE1 FIGURE
% -------------------------------------------------------------------------

@@

% -------------------------------------------------------------------------
% 
% -------------------------------------------------------------------------

In Obsidian, the same kernel (the local computation) is implemented as follows:

\noindent The {\tt vecAdd} program takes two {\tt SPull} input arrays and gives one 
as result. The Obsidian {\tt vecAdd} is more general than the CUDA version. From it 
we can generate CUDA kernels for any element type that supports the {\tt (+)} operation. 
In Haskell we know this is true for types that are member of the {\tt Num} typeclass.

There are some important differences between CUDA and Obsidian programs. One is that with Obsidian 
we generate kernels for fixed size local computations. This is to avoid doing out-of-bounds check 
which would mean conditionals in the code. Conditionals are problematic from a performance 
point of view. Another difference is that in Obsidian the programmer specifies the entire 
computation, including how it is mapped over the blocks of the GPU. This is done implicitly 
in CUDA, based on the indexing arithmetic used in the kernel and the number of blocks and 
threads provided at kernel launch. This means that the Obsidian programmer needs to 
write a program that performs this mapping:

\begin{small} 
\begin{Verbatim}[samepage=true]
vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
vecAddG i1 i2  =
  pConcat $
  zipWith vecAdd
          (splitUp 32 i1)
          (splitUp 32 i2)  
\end{Verbatim}
\end{small} % $

The {\tt vecAddG} program specifies a computation on large arrays; the large input arrays 
are split up into pieces that are operated upon by the local computation, {\tt vecAdd}, 
specified above. Each input array passed to this function must have length a multiple of 32.
Since the {\tt splitUp} operation splits the arrays into chunks of 32 elements, the body 
of the generated kernel will be constructed specifically for this size. {\tt zipWith} is 
used to map the local computation onto each chunk of the input arrays. The result is an 
array of arrays that is flattened with the {\tt pConcat} function. The CUDA code below 
is generated from the {\tt vecAddG} program:

\begin{small}
\begin{Verbatim}[samepage=true] 
extern "C" __global__ void vecAdd(int32_t* input0, uint32_t n0,
                                  int32_t* input1, uint32_t n1,
                                  int32_t* output2)
{
    uint32_t tid = threadIdx.x;

    if (blockIdx.x < min(n0 / 32U, n1 / 32U)) {
        output2[blockIdx.x * 32U + tid] = 
            input0[blockIdx.x * 32U + tid] +
            input1[blockIdx.x * 32U + tid];
    }
}
\end{Verbatim}
\end{small}

There are some superficial differences between the code generated using Obsidian and the 
handwritten example above. One such difference is that the index computation is not shared. 
Obsidian relies on the CUDA compiler to discover such sharing via a common subexpression 
elimination optimisation pass. Another difference is that Obsidian has inserted a conditional 
that only executes the body for certain blocks. Discussion of the causes for that conditional 
is postponed until a later section {\bf \emph{POINT TO SECTION}}. A real difference is that 
the kernel is generated for a fixed size, here there is a constant (32) where in the handwritten
code we used {\tt blockDim.x}. If those differences are set aside, the two kernels are identical. 
Figure~\ref{fig:OBSIDIANHOST1} shows how the {\tt vecAddG} kernel can be launched from 
within Haskell, using Obsidian's rudimentary support for launching GPU computations.

An alternative way to implement {\tt vecAddG} is to inline the local computation 
directly. This makes the entire program look like this.

\begin{small}
\begin{Verbatim}[samepage=true]
vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
vecAddG i1 i2  =
  pConcat $
  zipWith (zipWith (+)) 
          (splitUp 32 i1)
          (splitUp 32 i2)
\end{Verbatim} 
\end{small} % $

\noindent The generated code looks exactly as before.

% -------------------------------------------------------------------------
% OBSIDIANHOST1 FIGURE
% -------------------------------------------------------------------------

\begin{figure} 
\begin{small}
\begin{Verbatim}[samepage=true]
performVecAdd =
  withCUDA $
  do
    kern <- capture 32 (vecAddG :: DPull EInt32
                                 -> DPull EInt32
                                 -> DPush Grid EInt32)

    useVector (V.fromList [0..32::Int32]) $ \i1 ->
      useVector (V.fromList [0..32::Int32]) $ \i2 -> 
        allocaVector 32 $ \(o :: CUDAVector Int32) ->
        do
          fill o 0
          o <== (1,kern) <> i1 <> i2

          r <- peekCUDAVector o
          lift $ putStrLn $ show r
\end{Verbatim} 
\end{small}

\caption{Host code for launching a simple Obsidian kernel on the GPU. \\ 
 {\tt capture} compiles an Obsidian program into CUDA. At this point the 
  types of the Obsidian program need to be made concrete. The number, 32, used as 
 an argument to {\tt capture} specifies that generated code should be specialised 
 for 32 threads per block. }
\label{fig:OBSIDIANHOST1}
\end{figure}

A CUDA kernel that uses shared memory takes the form:

\begin{small}
\begin{Verbatim}[samepage=true]
__global__ void kernel(int32_t *i, int32_t *r) {

@@

\noindent A portion of the input array is stored into shared memory. A 
barrier synchronisation is used to ensure that each thread within the block 
has stored its value into shared memory (only necessary if threads will 
communicate across warp boundaries). After computing on the data locally, 
it is written back into global memory. This kernel sets up two aliases for 
thread identities. A thread's global identity is called {\tt gid} and its local 
identity is called {\tt tid}. The global ID is used to index into the large global 
array; indexing into local shared memory is done using the local ID.

In Obsidian, local computations can also use shared memory. The Obsidian arrays 
we have seen in the examples, {\tt Pull} and {\tt Push} arrays do not directly 
correspond to actual data in memory. Pull and push arrays, rather represent ways 
to compute arrays. More information about these array representations is available 
in section {\bf \emph{REF SECTION}}. The Obsidian programmer can store intermediate 
arrays in shared memory by using a {\tt force} operation. The 
{\tt force} operation computes an array and stores it into shared memory.

The following kernel multiplies each element in an array by two and then adds one:

\begin{small}
\begin{Verbatim}[samepage=true] 
kernel :: Num a => SPull a -> SPull a
kernel = fmap (+1) . fmap (*2)
\end{Verbatim}
\end{small}

\noindent This kernel computes its result without any storage of intermediate results. 
In the generated CUDA code below, it can be seen that the (*2) and (+1) operations 
are applied without any intermediate storage.

\begin{small}
\begin{Verbatim}[samepage=true]
extern "C" __global__ void kernel(int32_t* input0, uint32_t n0,
                                  int32_t* output1)
{
    uint32_t tid = threadIdx.x;

    if (blockIdx.x < n0 / 32U) {
        output1[blockIdx.x * 32U + tid] = 
          input0[blockIdx.x * 32U + tid] * 2 + 1;
    }
}
\end{Verbatim}
\end{small}

Storing an intermediate result between the two operations can be accomplished by 
changing the kernel as follows.

\begin{small}
\begin{Verbatim}[samepage=true]
kernelF :: (MemoryOps a, Num a)
         => SPull a -> BProgram (SPull a)
kernelF = liftM (fmap (+1)) . force . fmap (*2)
\end{Verbatim} 
\end{small}

\FloatBarrier 
\subsection{Writing efficient GPU Kernels} 
 \emph{now takes up both CUDA and Obsidian aspects at the same time, pref. with code examples}
